RVD2: An ultra-sensitive variant detection model for heterogeneous population sequencing

### 2013-09-06 doc/bioinformatics2013/RVD2.tex
Finished parameter initialization section and appendix.

Once we have obtained an estimate of the posterior distribution, we test the hypothesis that there is a sub-population of DNA molecules in the pool with a non-reference base at position $j, \forall j \in 1, \ldots, J$. First, we test the hypothesis that the error rate in the case data, $\mu_j^{\text{case}}$ is greater than the error rate in the control data, $\mu_j^{\text{control}}$. If so, then we test the hypothesis that the high error rate is due to excess reads with a particular non-reference base rather than a uniform distribution over all possible non-reference bases.

\subsection{Initialization}
The initial values of the model parameters and latent variables is obtained by a method-of-moments procedure. Essentially, we set the parameter value (population moment) equal to the sample moment. In some cases, the method-of-moments estimate is outside the parameter space. Then, we XXX.

The estimated error probability is calculated from the observed sample
\begin{equation}
 \theta_{ij}=\frac{r_{ij}}{n_{ij}}
\end{equation}
The estimate of parameter $\mu_j$ is obtained using the expected value of $\theta_{ij}$, while $M_j$ is estimated using the moment estimates for the variance of $\theta_{ij}$.
\begin{equation}
 \mu_j=E(\theta_{ij}|\mu_j,M_j)
\end{equation}
\begin{equation}
 M_j=\frac{\mu_j(1-\mu_j)}{Var(\theta_{ij}|\mu_j,M_j)}-1
\end{equation}

Likewise,the estimate of hyperparameters $\mu_o$ and $M_0$ are
\begin{equation}
 \mu_0=E(\mu_j|\mu_0,M_0)
\end{equation}
\begin{equation}
 M_0=\frac{\mu_0(1-\mu_0)}{Var(\mu_j|\mu_0,M_0)}-1
\end{equation}

@@

We are able to draw samples from $p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j)$ analytically because of the Bayesian conjugacy between the prior $p(\theta_{ji} | \mu_j, M_j) \thicksim \text{Beta}(\mu_j, M_j)$ and the likelihood $p(r_{ji} | n_{ji}, \theta_{ji}) \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$.

\subsection{Sampling from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$}

\subsection{Bayesian hypothesis test for high error rate}

@@

\section{Appendix}
\subsection{Moments derivation}
The probability density function of beta distribution, for $x\in[0,1]$, and shape real parameter $\alpha>1$ and $\beta>1$ is 
\begin{equation}
\begin{split}
  f(x;\alpha,\beta) &= \frac{x^{\alpha-1}(1-x)^{\beta-1}}{\int_0^1x^{\alpha-1}(1-x)^{\beta-1}dx}\\ &=\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}x^{\alpha-1}(1-x)^{\beta-1}\\ &=\frac{1}{B(\alpha,\beta)}x^{\alpha-1}(1-x)^{\beta-1}
\end{split}
\end{equation}
where $\Gamma(z)$ is the Gammma function, $B(\alpha,\beta)$ is the beta function, and
\begin{equation}
\begin{split}
 B(\alpha,\beta) &=\int_0^1x^{\alpha-1}(1-x)^{\beta-1}dx\\  &= \frac{\Gamma(\alpha)\Gamma(\beta)}{\Gamma(\alpha+\beta)}
\end{split}
\end{equation}

The expected value (mean) of a Beta distribution random variable X with parameter $\alpha$ and $\beta$ is 
\begin{equation}
\begin{split}
  E[X] &= \int_0^1xf(x;\alpha,\beta)dx\\ &= \int_0^1x\frac{x^{\alpha-1}(1-x)^{\beta-1}}{B(\alpha,\beta)}dx\\ &= \frac{\int_0^1x^{\alpha}(1-x)^{\beta-1}dx}{B(\alpha,\beta)}\\ &= \frac{B(\alpha+1,\beta)}{B(\alpha,\beta)}\\ &= \frac{\Gamma(\alpha+1)\Gamma(\beta)}{\Gamma(\alpha+\beta+1)}\cdot\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}\\ &= \frac{\alpha\Gamma(\alpha)\Gamma(\beta)}{(\alpha+\beta)\Gamma(\alpha+\beta)}\cdot\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}\\ &= \frac{\alpha}{\alpha+\beta}
\end{split}
\end{equation}
Note that Gamma function satisfies the functional equation
\begin{equation}
 \Gamma(z) = z\Gamma(z)
\end{equation}

The variance of a Beta distribution random variable X with parameter $\alpha$ and $\beta$ is 
\begin{equation}
\begin{split}
  var(X) &= E[(X-E(X))^2]\\ &= E(X^2)-[E(x)]^2\\ &=\int_0^1x^2\frac{x^{\alpha-1}(1-x)^{\beta-1}}{B(\alpha,\beta)}dx-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{\int_0^1x^{\alpha+1}(1-x)^{\beta-1}dx}{B(\alpha,\beta)}-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{B(\alpha+2,\beta)}{B(\alpha,\beta)}-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{\Gamma(\alpha+2)\Gamma(\beta)}{\Gamma(\alpha+\beta+2)}\cdot\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{(\alpha+1)\alpha\Gamma(\alpha)\Gamma(\beta)}{(\alpha+\beta+1)(\alpha+\beta)\Gamma(\alpha+\beta)}\cdot\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{(\alpha+1)\alpha}{(\alpha+\beta+1)(\alpha+\beta)}-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{\alpha\beta}{(\alpha+\beta+1)(\alpha+\beta)^2}
\end{split}
\end{equation}

Instead of the standard scale parameter $\alpha$ and $\beta$, we use the mean $\mu$ and sample-size paramenter $M$ for the Beta distribution in our hierarchical model.
\begin{equation}
\begin{split}
  f(x;\mu,M) &=\frac{\Gamma(M)}{\Gamma(M\mu)\Gamma(M(1-\mu))}x^{M\mu-1}(1-x)^{M(1-\mu)-1}\\ &= \frac{1}{B(M\mu,M(1-\mu))}x^{M\mu-1}(1-x)^{M(1-\mu)-1}
\end{split}
\end{equation}
where 
\begin{equation}
 \mu = \frac{\alpha}{\alpha+\beta}
\end{equation}
\begin{equation}
 M = \alpha+\beta
\end{equation}
From Equation 24 and Equation 26 it can be seen that
\begin{equation}
 E(x|\mu,M) = \mu
\end{equation}
\begin{equation}
 Var(x|\mu,M) = \frac{\mu(1-\mu)}{M+1}
\end{equation}

>>>>

Once we have obtained an estimate of the posterior distribution for the case and control data, we use a two-stage test for variant at each position. First, we test whether the error rate in the case sample is significantly greater than the error rate in the control data for each position. Then, we test the hypothesis that the error rate is high is due to excess reads with a particular non-reference base. We accomplish the first test with a Bayesian posterior density test and the second with a $\chi^2$ goodness-of-fit test to a uniform distribution. In the remainder of this section, we describe the initialization, MCMC sampling approach and hypothesis testing procedure.

%%%%%%%%%%%%%
% Initialization
%%%%%%%%%%%%%
\subsection{Initialization}
The initial values for the model parameters and latent variables is obtained by a method-of-moments(MoM) procedure. MoM works by setting the population moment equal to the sample moment. A system of equations is formed such that the number of moment equations is equal to the number of unknown parameters and the equations are solved simultaneously to give the parameter estimates. We simply start with the data matrices $r$ and $n$ and work up the hierarchy of the graphical model solving for the parameters of each conditional distribution in turn.

We present the initial parameter estimates here and provide the derivations in Appendix~\ref{sec:appendix_mom}. The MoM estimate for replicate-level parameters are $\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}$. The estimates for the position-level parameters are $\tilde{\mu}_j = \frac{1}{N} \sum_{i=1}^N \theta_{ji}$ and $\tilde{M_j} = \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1$. The estimates for the genome-level parameters are $\tilde{\mu}_0 = \frac{1}{J} \sum_{j=1}^J \mu_j$ and $\tilde{M}_0 = \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1$.

The method-of-moments is known to have the potential to give estimates that are outside the true parameter bounds. For example, the MoM estimate $\tilde{M}_j$ may be negative while $M_j$ must be non-negative. We have not encounter such pathological behavior of the estimates in this application.

@@

Samples from the posterior distribution $p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j)$ are drawn analytically because of the Bayesian conjugacy between the prior $p(\theta_{ji} | \mu_j, M_j) \thicksim \text{Beta}(\mu_j, M_j)$ and the likelihood $p(r_{ji} | n_{ji}, \theta_{ji}) \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$. The posterior distribution is

\begin{equation}
	p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j) \thicksim \text{Beta}\left( \frac{r_{ji} + M_j \mu_j}{n_{ji} + M_j} , n_{ji} + M_j\right),
\end{equation}
where we have used the mean-precision parameterization of the beta distribution.

\subsection{Sampling from $p \left( \mu_j |\theta_{ji},\mu_0,M_0\right)$}

The posterior distribution for $\mu_j$ is
\begin{equation}
	p \left( \mu_j |\theta_{ij},\mu_0,M_0\right) = 
\end{equation}
\subsection{Bayesian hypothesis test for high error rate}

@@

\appendix
\section{Parameter Initialization}\label{sec:appendix_mom}
Since $r_{ji} \thicksim \text{Binomial}(n_{ji}, \theta_{ji})$, the first population moment is  $E[r_{ji}] = \theta_{ji} n_{ji}$ and the first sample moment is simply $m_1 = r_{ji}$. Therefore the MoM estimator is 
\begin{equation}
	\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}
\end{equation}

We take the MoM estimate, $\tilde{\theta}_{ji}$, as data for the next conditional distribution in the hierarchical model. The distribution is $\theta_{ji} \thicksim \text{Beta}(\mu_jM_j, (1-\mu_j)M_j)$. The first and second population moments are
\begin{eqnarray}
	E[\theta_{ji}] =& \mu_j,\\
	\text{Var}[\theta_{ji}] =& \frac{\mu_j(1-\mu_j)} { M_j + 1 }.
\end{eqnarray}
The first and second sample moments are $m_1 = \frac{1}{N}\sum_{i=1}^N \theta_{ji}$ and $m_2 = \frac{1}{N}\sum_{i=1}^N \theta_{ji}^2$. Setting the population moments equal to the sample moments and solving for $\mu_j$ and $M_j$ gives
\begin{eqnarray}
	\tilde{\mu}_j =& \frac{1}{N} \sum_{i=1}^N \theta_{ji}, \\
	\tilde{M_j} =& \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1.
\end{eqnarray}

Following the same procedure for the parameters of $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$ gives the following MoM estimates
\begin{eqnarray}
	\tilde{\mu}_0 =& \frac{1}{J} \sum_{j=1}^J \mu_j \\
	\tilde{M}_0 =& \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1.
\end{eqnarray}
